Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-05-08 manuscript.Rmd
Final tidying; LAI in figures replaced with L for consistency

# Number of months of study
r <- range(df$Date)
nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance to net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

@@

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig. SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig. S2). We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig. S2), giving *k* = `r kopt`. These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

\
The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXXXX), published on figshare (doi:XXXX). The repository includes the processed data needed to run the analyses and figures (*raw data needs to be shared too, have to figure out how though!*).

@@

\
An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see [@Hall2015]). This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower $\tau_d$ than before the insect attack.

@@

\
The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

\
We estimated gross leaf production from measured litter production ($P_L$) and temporal dynamics of $L_\tau$. Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to $P_L$ to arrive at an estimate of the leaf production. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and $P_L$. We found no significant difference in either litter or leaf production between C~a~ treatments (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related $P_L$ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig. 6a). If leaf lifespan is constant, $P_L$ would be linearly related to $L_\tau$. Instead, we found that in eC~a~, $P_L$ was lower at a given $L_\tau$ as compared to aC~a~. We estimated mean leaf lifespan from Fig. 6a as the ratio of $L_\tau$ to annual $P_L$. We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig. 6b), with LL = `r LLtest$estimate1`yr in aC~a~ and `r LLtest$estimate2`yr in eC~a~.

@@

```{r}
options(scipen = 1, digits = 5)
```

@@

**Figure 2**. Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA). This relationship was used to calculate the *L* anomaly in Fig. 3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.

>>>>

# Number of months of study
r <- range(df$Date)
r[2] <- maxdate
nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

@@

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig. SI2, R^2^=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig. S2). We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig. S2), giving *k* = `r kopt`. These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

\
The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXXXX), published on figshare (doi:XXXX). The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, with the rmarkdown package in Rstudio, which uses Pandoc for conversion).

@@

\
An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015]. This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower $\tau_d$ than before the insect attack.

@@

\
The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$std.error[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

\
We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $L_\tau$. Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to P~L~ to arrive at an estimate of the leaf production. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and P~L~. We found no significant difference in either litter or leaf production between C~a~ treatments (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related P~L~ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig. 6a). If leaf lifespan is constant, P~L~ would be linearly related to $L_\tau$. Instead, we found that in eC~a~, P~L~ was lower at a given $L_\tau$ as compared to aC~a~. We estimated mean leaf lifespan from Fig. 6a as the ratio of $L_\tau$ to annual total P~L~. We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig. 6b), with LL = `r LLtest$estimate1`yr in aC~a~ and `r LLtest$estimate2`yr in eC~a~.

@@

```{r, echo=FALSE, message=FALSE}
options(scipen = 1, digits = 2)
```

@@

**Figure 2**. Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA). This relationship was used to calculate the *L* anomaly in Fig. 3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.
